Circumpolar variations in Southern Ocean eddy dynamics: insights from an ensemble modelling approach

### 2021-09-27 paper/main.tex
fix Sally affiliation
don't enforce bibtex
adds Sally's remakrs + cite NCC paper + punctuate all eqs

@@

\authors{
Andrew McC. Hogg\affil{1}, Thierry Penduff\affil{2}, Sally E. Close\affil{3}, William K. Dewar\affil{4},\\Navid C. Constantinou\affil{1}\quad and\quad Josu{\'e} Mart{\'i}nez Moreno\affil{1}
}

\affiliation{1}{Research School of Earth Sciences and ARC Centre of Excellence for Climate Extremes,\\ Australian National University, Canberra, Australia}
\affiliation{2}{Universit{\'e} Grenoble Alpes, CNRS, IRD, Grenoble-INP, Institut des G{\'e}osciences de l'Environnement (IGE), Grenoble, France}
\affiliation{3}{Institut des Géosciences de l'Environnement, CNRS/Univ. Grenoble Alpes/G-INP/IRD, France}
\affiliation{3}{Florida State University, Tallahassee, FL, USA}

@@

The strength of the Southern Ocean eddy field is usually characterised by extracting the kinetic energy of transient motions; referred to by oceanographers as eddy kinetic energy (EKE). 
It is important to highlight that EKE includes all transient motion, not just coherent vortices \citep{Martinez-Moreno2019}, and thus care needs to be taken in interpreting this metric.
Eddy kinetic energy has a complex relationship with the forces that drive the ocean circulation.
\citet{Meredith-Hogg-2006} found significant variations in the area-averaged eddy field in some regions, and argued for a 2-3 year lag between events in wind stress forcing and EKE anomalies.
\citet{Patara2016}, using a realistic high-resolution ocean model found that EKE does have a lagged response to wind stress anomalies, but that this relationship is variable around the Southern Ocean.
Idealised models over a wide range of parameter space \citep{Sinha2016} have highlighted that the timescale of the perturbation is critical in determining the EKE response, with shorter perturbations having a faster, Ekman-related response.
Thus, current knowledge suggests that there is a relationship between wind forcing and the Southern Ocean, but that the nature of this relationship, and the role of intrinsic variability, need to be clarified.\footnote{{\color{red}Thierry: the notion of intrinsic variability could be introduced a bit more before this, explaining for instance that it emerges frrom oceanic non-linearities under constant or seasonal forcing and persists under variable forcing (Leroux et al 2018), it has a random phase (``chaotic'' character), it mostly emerges at mesoscale and can cascade toward interannual to decadal time scales and O(1000km) space scales (Serazin et al 2018).}}

On longer timescales it has also been proposed that EKE has increased over recent decades \citep{Hogg2015, Martinez-Moreno2019}.
The robustness of this wind-EKE relationship in the Southern Ocean was recently investigated by \citet{Zhang2021}, who used crossover data from satellite observations \citep[as in][]{Hogg2015} to better estimate the EKE on regional scales. 
When fine graining these calculations it was found that only a single region (30$^\circ$ in longitude) expressed a significant long-term trend in EKE.
This finding  suggests that previous characterisations of the response may have been dominated by a small number of regional events, and that at other times the intrinsic variance may dominate the signal.

Thus, despite the proposed importance of the Southern Ocean eddy field to ocean circulation, the nature of the EKE variability remains unclear.
A primary complication is what can be inferred from an admittedly short satellite record; in particular, whether individual events can be attributed to forcing changes, or to intrinsic variability, or a combination of the two.
In this paper, we address this question by examining the intrinsic variability of the Southern Ocean eddy field in an ensemble of eddy-permitting ocean simulations.
We use the OCCIPUT ensemmble of global ocean/sea-ice hindcast simulations \citep{Penduff-etal-2014, Leroux2018}, a 50-member ensemble of hindcast simulations.
We examine both the intrinsic variance of the eddy field, and extract the ``forced'' (ensemble mean) component of the variability.
This variability is examined on a circumpolar and regional basis, to better understand the regional differences and processes which contribute to the eddy field.

@@

The methodology employed in this study is derived from the probabilistic approach to ocean modelling outlined by \citet{Bessieres2017}.
We use output from the OCCIPUT (OceaniC Chaos – ImPacts, strUcture, predicTability) ensemble of 50 eddy-permitting global ocean--sea ice simulations, based on the ORCA025 implementation \citep[e.g.][]{Barnier2006} of the NEMO modelling system \citep{Madec2012}. 
The model grid has a 1/4$^\circ$ horizontal resolution and 75 geopotential levels. We employ the probabilistic ocean modelling approach outlined by \citet{Bessieres2017}: \emph{(i)}~the 50 ensemble members are initialized in 1960 from a one-member 21-year spinup; \emph{(ii)}~the ensemble spread is seeded by activating a stochastic perturbation \citep{Brankart-etal-2015} during one year in 1960, which is shut off afterwards; \emph{(iii)}~the 50 ensemble members are driven between 1960 and 2015 by the same realistic atmospheric forcing function based on the ERA-40 and ERA-Interim reanalyses (Drakkar Forcing Set DFS5.2; \citet{Dussin-etal-2016}); the wind stress is computed without ocean current feedbacks, hence ensuring that the same momentum fluxes are applied to all members. 
We focus our analyses on the period 1980-2015, thus yielding an effective spinup duration of 41 years in each member.

\subsection{Estimating geostrophic eddy velocity}
For each ensemble member, the sea surface height, $\eta_i$, is saved as a 5-daily average (where $i$ represents the ensemble member) . 
The sea level anomaly, $\eta_i'$, is calculated as the transient component of sea level, given by 
\begin{linenomath*}
\begin{equation}
\eta_i' = \eta_i - \overline{\eta_i}
\end{equation}
\end{linenomath*}
where $\overline{\cdot}$ represents the temporal average over the simulation.
The sea level anomaly is detrended, and then used to calculate the eddy velocity field.
Surface eddy velocity in each ensemble member, $\mathbf{u}_i' = (u_i', v_i')$ is estimated from the detrended sea level anomaly via the geostrophic relation,
\begin{linenomath*}
\begin{equation}
u_i' = - \frac{g}{f} \frac{\partial \eta_i'}{\partial y}; \quad v_i' = \frac{g}{f} \frac{\partial \eta_i'}{\partial x}
\end{equation}
\end{linenomath*}
where $g$ is the acceleration due to gravity and $f$ is the local Coriolis parameter.
Anomalous values of $\mathbf{u}_i'$ close to coastlines are removed.

@@

The surface eddy kinetic energy for each ensemble member is calculated as 
\begin{linenomath*}
\begin{equation}
E_i = \frac{1}{2}(u_i'^2 + v_i'^2)
\end{equation}
\end{linenomath*}
and subsampled at monthly temporal resolution.
It is notable that $E_i$ contains a large component of seasonal variation \citep{Martinez-Moreno2021} which dominates the statistics.
To look at the forced response, we therefore deseasonalise $E_i$, by removing the climatological mean for each month.
We then proceed to compute ensemble statistics from this deseasonalised EKE.

@@

The eddy kinetic energy can be averaged over the $N$ ensemble members to give the ensemble mean EKE, written as
\begin{linenomath*}
\begin{equation}
\langle E_i \rangle = \frac{1}{N} \sum_{i=1}^N E_i.
\end{equation}
\end{linenomath*}
The argument can be made that, since each ensemble member is forced with identical atmospheric conditions, the forced EKE response is encapsulated by this ensemble mean quantity.
On the other hand, the intrinsic variability of EKE in each ensemble member is found by taking the difference from the ensemble mean for each member,\footnote{{\color{red}Thierry: instead of using the prime exponent (used for temporal anomalies), I suggest using e.g. a star (*) exponent for ensemble anomalies.}}
\begin{linenomath*}
\begin{equation}
E_i' = E_i - \langle E_i \rangle.
\end{equation}
\end{linenomath*}
With these expressions in hand, we follow \citet{Leroux2018} to define the  time-variance of the ensemble-mean EKE,
\begin{linenomath*}
\begin{equation}
\sigma^2_{\langle E \rangle} = \frac{1}{T} \sum_{t=1}^T \left(\langle E_i \rangle -  \overline{\langle E_i \rangle}\right)^2,
\end{equation}
\end{linenomath*}
which represents the variance of the forced eddy response.
Analogously, the intrinsic variance emerges from the variance,
\begin{linenomath*}
\begin{equation}
\epsilon^2(t) = \frac{1}{N} \sum_{i=1}^N E_i'(t)^2
\end{equation}
\end{linenomath*}
indicating the spread of each member from the ensemble mean.
The fraction of intrinsic variance is then computed as the ratio of the intrinsic variance to the total variance,
\begin{linenomath*}
\begin{equation}
R_i =  \frac{\overline{\epsilon^2}}{\overline{\epsilon^2} + \sigma^2_{\langle E \rangle}}.
\end{equation}
\end{linenomath*}
When $R_i$ approaches unity, the system is dominated by intrinsic variability, and at the limit  of $R_i \to 0$ the system is solely responding to forced variability induced by atmospheric forcing.

To examine the relationship between the EKE (of either the ensemble mean, or from individual ensemble members) we are guided by previous work which emphasises the role of wind stress in governing the eddy variability at different timescales \citep[e.g.][]{Hogg2015, Sinha2016}.
Thus, we look at the correlation coefficient, $r$, at a range of lags between the wind stress $\tau$ (which is identical for all ensemble members) and EKE.
For each correlation we evaluate the statistical significance of the correlation \citep[following, e.g.][]{Santer2000}.
We first calculate the effective sample size, $N_e$, where
\begin{linenomath*}
\begin{equation}
N_e \equiv N \frac{1-r_1 r_2}{1+r_1 r_2}
\end{equation}
\end{linenomath*}
where $N$ is the total number of samples (444 months for this timeseries), $r_1$ is the lag-1 autocorrelation for EKE and $r_2$ is the lag-2 autocorrelation for wind stress.
We use the Students t-test to infer statistical significance (at the 95\%-level) based on this effective sample size, when
\begin{linenomath*}
\begin{equation}
T = \frac{r \sqrt{N_e}}{\sqrt{1-r^2}} > 2.
\end{equation}
\end{linenomath*}
Lagged correlation estimates indicate when this significance test is satisfied.

@@

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\hsize]{Figure1}
\caption{(a) Snapshot (5-day average) from a single ensemble member showing Southern Ocean EKE, $E_i$, and (b) time-averaged ensemble mean EKE, $\overline{\langle E_i\rangle}$.  }
\label{Fig:1}
\end{center}
\end{figure}

>>>>

% !TEX program = pdflatex
% !TEX encoding = UTF-8 Unicode

@@

\authors{
Andrew McC. Hogg\affil{1}, Thierry Penduff\affil{2}, Sally E. Close\affil{2}, William K. Dewar\affil{3},\\Navid C. Constantinou\affil{1}, \quad and\quad Josu{\'e} Mart{\'i}nez Moreno\affil{1}
}

\affiliation{1}{Research School of Earth Sciences and ARC Centre of Excellence for Climate Extremes,\\ Australian National University, Canberra, Australia}
\affiliation{2}{Universit{\'e} Grenoble Alpes, CNRS, IRD, Grenoble-INP, Institut des G{\'e}osciences de l'Environnement (IGE), Grenoble, France}
\affiliation{3}{Florida State University, Tallahassee, FL, USA}

@@

The strength of the Southern Ocean eddy field is usually characterised by extracting the kinetic energy of transient motions; referred to by oceanographers as eddy kinetic energy (EKE). 
It is important to highlight that EKE includes all transient motion, not just coherent vortices \citep{Martinez-Moreno2019}, and thus care needs to be taken in interpreting this metric.
Eddy kinetic energy has a complex relationship with the forces that drive the ocean circulation.
\citet{Meredith-Hogg-2006} found significant variations in the area-averaged eddy field in some regions, and argued for a 2-3 year lag between events in wind stress forcing and EKE anomalies.
\citet{Patara2016}, using a realistic high-resolution ocean model found that EKE does have a lagged response to wind stress anomalies, but that this relationship is variable around the Southern Ocean.
Idealised models over a wide range of parameter space \citep{Sinha2016} have highlighted that the timescale of the perturbation is critical in determining the EKE response, with shorter perturbations having a faster, Ekman-related response.
Thus, current knowledge suggests that there is a relationship between wind forcing and the Southern Ocean, but that the nature of this relationship, and the role of intrinsic variability, need to be clarified.\footnote{{\color{red}Thierry: the notion of intrinsic variability could be introduced a bit more before this, explaining for instance that it emerges frrom oceanic non-linearities under constant or seasonal forcing and persists under variable forcing \citep{Leroux2018}, it has a random phase (``chaotic'' character), it mostly emerges at mesoscale and can cascade toward interannual to decadal time scales and O(1000$\,$km) space scales \cite{Serazin-etal-2018}.}}

On longer timescales it has also been proposed that EKE has increased over recent decades \citep{Hogg2015, Martinez-Moreno2019, Martinez-Moreno2021-ncc}.
The robustness of this wind-EKE relationship in the Southern Ocean was recently investigated by \citet{Zhang2021}, who used crossover data from satellite observations \citep[as in][]{Hogg2015} to better estimate the EKE on regional scales.
When fine graining these calculations it was found that only a single region (30$^\circ$ in longitude) expressed a significant long-term trend in EKE.
This finding  suggests that previous characterisations of the response may have been dominated by a small number of regional events, and that at other times the intrinsic variance may dominate the signal.

Thus, despite the proposed importance of the Southern Ocean eddy field to ocean circulation, the nature of the EKE variability remains unclear.
A primary complication is what can be inferred from an admittedly short satellite record; in particular, whether individual events can be attributed to forcing changes, or to intrinsic variability, or a combination of the two.
In this paper, we address this question by examining the intrinsic variability of the Southern Ocean eddy field in an ensemble of eddy-permitting ocean simulations.
We use the OCCIPUT ensemble of global ocean/sea-ice hindcast simulations \citep{Penduff-etal-2014, Leroux2018}, a 50-member ensemble of hindcast simulations\footnote{{\color{blue}Sally: The definition of OCCIPUT perhaps need to be moved up here from the paragraph below?}}.
We examine both the intrinsic variance of the eddy field, and extract the ``forced'' (ensemble mean) component of the variability.
This variability is examined on a circumpolar and regional basis, to better understand the regional differences and processes which contribute to the eddy field.

@@

The methodology employed in this study is derived from the probabilistic approach to ocean modelling outlined by \citet{Bessieres2017}.
We use output from the OCCIPUT (OceaniC Chaos – ImPacts, strUcture, predicTability) ensemble of 50 eddy-permitting global ocean--sea ice simulations, based on the ORCA025 implementation \citep[e.g.][]{Barnier2006} \footnote{{\color{blue}Sally: I think there's a problem in the bibtex file here: the citation seems to be coming up as Bernard instead of Barnier}}of the NEMO modelling system \citep{Madec2012}. 
The model grid has a 1/4$^\circ$ horizontal resolution and 75 geopotential levels. We employ the probabilistic ocean modelling approach outlined by \citet{Bessieres2017}: \emph{(i)}~the 50 ensemble members are initialized in 1960 from a one-member 21-year spinup; \emph{(ii)}~the ensemble spread is seeded by activating a stochastic perturbation \citep{Brankart-etal-2015} during one year in 1960, which is shut off afterwards; \emph{(iii)}~the 50 ensemble members are driven between 1960 and 2015 by the same realistic atmospheric forcing function based on the ERA-40 and ERA-Interim reanalyses (Drakkar Forcing Set DFS5.2; \citet{Dussin-etal-2016}); the wind stress is computed without ocean current feedbacks, hence ensuring that the same momentum fluxes are applied to all members. 
We focus our analyses on the period 1980-2015, thus yielding an effective spinup duration of 41 years in each member.

\subsection{Estimating geostrophic eddy velocity}
For each ensemble member, the sea surface height, $\eta_i$, is saved as a 5-daily average (where $i$ represents the ensemble member) . {\color{blue} Sally: For each ensemble member, the member's global mean sea level anomaly is subtracted from all grid points at every time step to correct for spurious terms introduced by the use of the Boussinesq approximation \citep{Greatbatch1994}. The model drift, which is potentially nonlinear, is then corrected for by detrending using a LOWESS filter \citep{Cleveland1979} in combination with a 5th-order spline; full details of this preprocessing may be found in \citep{Close2020}.}

The sea level anomaly, $\eta_i'$, is calculated as the transient component of sea level, given by 
\begin{linenomath*}
\begin{equation}
\eta_i' = \eta_i - \overline{\eta_i},
\end{equation}
\end{linenomath*}
where $\overline{\cdot}$ represents the temporal average\footnote{{\color{blue}Sally: see above: instead of the temporal average it's the nonlinear trend that's removed}} over the simulation.
The sea level anomaly is detrended, and then used to calculate the eddy velocity field.
Surface eddy velocity in each ensemble member, $\mathbf{u}_i' = (u_i', v_i')$ is estimated from the detrended sea level anomaly via the geostrophic relation,
\begin{linenomath*}
\begin{equation}
u_i' = - \frac{g}{f} \frac{\partial \eta_i'}{\partial y} \quad \text{and} \quad v_i' = \frac{g}{f} \frac{\partial \eta_i'}{\partial x},
\end{equation}
\end{linenomath*}
where $g$ is the acceleration due to gravity and $f$ is the local Coriolis parameter.
Anomalous values of $\mathbf{u}_i'$ close to coastlines are removed.

@@

The surface eddy kinetic energy for each ensemble member is calculated as 
\begin{linenomath*}
\begin{equation}
E_i = \frac{1}{2}(u_i'^2 + v_i'^2),
\end{equation}
\end{linenomath*}
and subsampled at monthly temporal resolution.
It is notable that $E_i$ contains a large component of seasonal variation \citep{Martinez-Moreno2021} which dominates the statistics.
To look at the forced response, we therefore deseasonalise $E_i$, by removing the climatological mean for each month.
We then proceed to compute ensemble statistics from this deseasonalised EKE.

@@

The eddy kinetic energy can be averaged over the $N$ ensemble members to give the ensemble mean EKE, written as
\begin{linenomath*}
\begin{equation}
\langle E_i \rangle = \frac{1}{N} \sum_{i=1}^N E_i.
\end{equation}
\end{linenomath*}
The argument can be made that, since each ensemble member is forced with identical atmospheric conditions, the forced EKE response is encapsulated by this ensemble mean quantity.
On the other hand, the intrinsic variability of EKE in each ensemble member is found by taking the difference from the ensemble mean for each member,\footnote{{\color{red}Thierry: instead of using the prime exponent (used for temporal anomalies), I suggest using e.g. a star (*) exponent for ensemble anomalies.}}
\begin{linenomath*}
\begin{equation}
E_i' = E_i - \langle E_i \rangle.
\end{equation}
\end{linenomath*}
With these expressions in hand, we follow \citet{Leroux2018} to define the  time-variance of the ensemble-mean EKE,
\begin{linenomath*}
\begin{equation}
\sigma^2_{\langle E \rangle} = \frac{1}{T} \sum_{t=1}^T \left(\langle E_i \rangle -  \overline{\langle E_i \rangle}\right)^2,
\end{equation}
\end{linenomath*}
which represents the variance of the forced eddy response.
Analogously, the intrinsic variance emerges from the variance,
\begin{linenomath*}
\begin{equation}
\epsilon^2(t) = \frac{1}{N} \sum_{i=1}^N E_i'(t)^2,
\end{equation}
\end{linenomath*}
indicating the spread of each member from the ensemble mean.
The fraction of intrinsic variance is then computed as the ratio of the intrinsic variance to the total variance,
\begin{linenomath*}
\begin{equation}
R_i =  \frac{\overline{\epsilon^2}}{\overline{\epsilon^2} + \sigma^2_{\langle E \rangle}}.
\end{equation}
\end{linenomath*}
When $R_i$ approaches unity, the system is dominated by intrinsic variability, and at the limit  of $R_i \to 0$ the system is solely responding to forced variability induced by atmospheric forcing.

To examine the relationship between the EKE (of either the ensemble mean, or from individual ensemble members) we are guided by previous work which emphasises the role of wind stress in governing the eddy variability at different timescales \citep[e.g.][]{Hogg2015, Sinha2016}.
Thus, we look at the correlation coefficient, $r$, at a range of lags between the wind stress $\tau$ (which is identical for all ensemble members) and EKE.
For each correlation we evaluate the statistical significance of the correlation \citep[following, e.g.][]{Santer2000}.
We first calculate the effective sample size, $N_e$, where
\begin{linenomath*}
\begin{equation}
N_e \equiv N \frac{1-r_1 r_2}{1+r_1 r_2},
\end{equation}
\end{linenomath*}
where $N$ is the total number of samples (444 months for this timeseries), $r_1$ is the lag-1 autocorrelation for EKE and $r_2$ is the lag-2 autocorrelation for wind stress.
We use the Students t-test to infer statistical significance (at the 95\%-level) based on this effective sample size, when
\begin{linenomath*}
\begin{equation}
T = \frac{r \sqrt{N_e}}{\sqrt{1-r^2}} > 2.
\end{equation}
\end{linenomath*}
Lagged correlation estimates indicate when this significance test is satisfied.

@@

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\hsize]{Figure1}
\caption{(a) Snapshot (5-day average) from a single ensemble member showing Southern Ocean EKE, $E_i$, and (b) time-averaged ensemble mean EKE, $\overline{\langle E_i\rangle}$.}
\label{Fig:1}
\end{center}
\end{figure}
